---
title: "Lab 09: Algorithmic Bias"
author: "Cailey Fay"
date: "3.17.26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(readr)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15)

glimpse(compas)
```

## Exercise 1 
There are `r nrow(compas)` rows and `r ncol(compas)` columns of the frame. Each row represents a person. 

## Exercise 2 
There are `r count(distinct(compas, name))` unique people in the frame. This is not the same as the total number of observations, perhaps because some repeat offenders get listed twice. 

## Exercise 3
The distribution is quite flat, but concentrated at the lower scores. 
```{r}
hist(compas$decile_score)
```
## Exercise 4

```{r}
ggplot(compas, mapping = aes(x=fct_infreq(race), fill = race)) +
  geom_bar() +
 # coord_polar(theta = "y") +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(title = "Distribution of Race",
       y= "Frequency",
       x = "Race") +
  scale_fill_brewer()

ggplot(compas, mapping = aes(x=sex, fill = sex)) +
  geom_bar() +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(title = "Defendant Sex",
       y= "Frequency",
       x = "Sex") +
  scale_fill_brewer()

ggplot(compas, mapping = aes(x=age_cat, fill = age_cat)) +
  geom_bar() +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(title = "Defendant Age",
       y= "Frequency",
       x = "Age Category") +
  scale_fill_brewer()
```

## Excercise 4 Stretch
I might revisit this, but right now I want to invest my time elsewhere. 
```{#r}
ggplot(compas, mapping = aes(x=age_cat, fill = age_cat)) +
  geom_bar() +
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(title = "Defendant Age",
       y= "Frequency",
       x = "Age Category") +
  scale_fill_brewer() +
  facet_grid(~race, age_cat, race)

```

## Exercise 5
They do correspond with actual recidivism, but not remarkably so. This chart isn't amazing but it gets my point across. 0 = no reoffense, 1 = reoffense. 
```{r}
compas <- compas %>%
  mutate(nom_decile_score = as.factor(decile_score))


ggplot(compas, mapping = aes(x = nom_decile_score, fill = as.factor(two_year_recid))) +
  geom_bar(position = "fill") +
  scale_fill_brewer() +
  theme_minimal() + 
  labs(title = "Proportion of Actual Recidivism by Risk Class",
                    x = "Risk Decile",
                    y = "Proportion of Recidivism (or Lackthereof)",
          fill = "Recidivism")

#I am going to use regression to see if the algorithm significantly predicts recidivism. It does, but it only explains 12% of the variance in recidivism.  
reg <- lm(two_year_recid ~ decile_score, compas)
summary(reg)
```

## Exercise 6 & 7 

Overall, the algorithm was right 4,670 times out of 7,214, which is 64% accurate. 

```{r}
compas <- compas %>%
  mutate(compas_correct_or_not = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ 1, #True positive
    decile_score < 7 & two_year_recid == 0 ~ 1, #True negative
    decile_score >=7 & two_year_recid == 0 ~ 0, #False positive
    decile_score < 7 & two_year_recid == 1 ~ 0)) #False negative 

compas <- compas %>%
  mutate(compas_class = as.factor(case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP", #True positive
    decile_score <=4 & two_year_recid == 0 ~ "TN", #True negative
    decile_score >=7 & two_year_recid == 0 ~ "FP", #False positive
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    decile_score <4 & decile_score < 7 ~ "Mid"))) #False negative 

table(compas$compas_class)
table(compas$compas_correct_or_not)

#these are ugly, just for my own understanding
ggplot(compas, aes(x=compas_class)) +
  geom_bar()

ggplot(compas, aes(x=compas_correct_or_not, fill = compas_class)) +
  geom_bar(position = "fill")
```

Additional Exercises  
Now let’s assess how well the COMPAS algorithm performs across different demographic groups. Like ProPublica, we’ll focus on race, but you can explore others.

## Exercise 8 
Black defendants are more frequently assigned as high risk, whereas white defandants tend to have lower risk scores. 
```{r}
compas %>%
  filter(race == "African-American" | race == "Caucasian") %>%
ggplot(aes(x=nom_decile_score)) +
  geom_bar() +
  facet_wrap(~race) +
  labs(title = "Distribution of Risk Decile Scores by Race",
       x="Decile Score",
       y = "Frequency")

```

## Exercise 9 
 38.5% of black defendants are high risk versus 16.5% of white defendants. This is a large dispairity. 
```{r}
Bcompas <- compas %>%
  filter(race == "African-American")

Wcompas <- compas %>%
  filter(race == "Caucasian")

table(Bcompas$decile_score)
table(Wcompas$decile_score)
```

## Exercise 10 
The algorithm tends to incorrectly assign white defendants as low risk and black defendants as high risk. 

The proportion of non-recidivists who had false positives is much higher for black defendants. 
```{r}
#25% false positives for black defendants 
Bnon_recidivists <- Bcompas %>%
  filter(two_year_recid == 0)
table(Bnon_recidivists$compas_class)

#9% false positives for white defendants 
Wnon_recidivists <- Wcompas %>%
  filter(two_year_recid == 0)
table(Wnon_recidivists$compas_class)
```

The proportion of recidivists who had false negatives was much higher for white defendants.  
```{r}
#30% false negatives for black defendants 
Brecidivists <- Bcompas %>%
  filter(two_year_recid == 1)
table(Brecidivists$compas_class)

#48% false negatives for white defendants 
Wrecidivists <- Wcompas %>%
  filter(two_year_recid == 1)
table(Wrecidivists$compas_class)
```

## Exercise 11 
False positives were twice as common for the black defendants versus white. (Black defendant FP is 12%
White defendant FP is 6%)
```{r}
table(Bcompas$compas_class)
table(Wcompas$compas_class)

compas %>%
    filter(race == "African-American" | race == "Caucasian") %>%
ggplot(aes(x=race, fill = compas_class)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("FP" = "red","FN" = "orange", "TP" = "grey", "TN" = "grey", "NA" = "grey")) +
  theme_minimal() +
  labs(title = "Proportion of Risk Classifications Among Black and White Defendants",
       x = "Race",
       y = "Frequency",
       fill = "Risk Class")
```
## Exercise 12 

The algorithm weighs prior convictions more heavily for black defendants. The colors won't play nice in my graph, so thats why its faceted instead. 
```{r}
compas %>%
    filter(race == "African-American" | race == "Caucasian") %>%
ggplot(aes(x=decile_score, y=priors_count)) +
  geom_jitter(alpha = 0.7) + 
 # scale_color_manual(values = c("African-American" = "blue", "Caucasion"="midnightblue")) +
  facet_wrap(~race) +
  theme(legend.position = "none") +
  theme_minimal() +
  labs(title = "Association of Prior Convictions and Risk Scores, by Race",
       x = "Risk Score",
       y = "Prior Convictions")
#I can't get the colors to play nice 
```
## Exercise 13

I don't think the algorithm is fair. If it were, the false positive and false negative ratings would be more similar across races. It is quite alarming that the algorithm is more likely to incorrectly conclude that a white person isn't going to re offend or that a black person will. I would be interested in seeing what these rates are for the other races in the dataset. 

My visualization of the calibration actually makes it seem like maybe the algorithm is more fair than I originally thought (the second one, not the first). This chart is in line with Northpointe's claim. Similar proportions of defendants at each decile reoffended. However, white people are more likely to be given a lower risk score, so my chart was probably biased from the beginning. The classification system is biased, so I am personally not sure what to think. 

```{r}
# To check calibration, you can calculate recidivism rates by score for each race
df<- compas %>%
  filter(race == "African-American" | race == "Caucasian") %>%
  group_by(nom_decile_score, race) %>%
  summarize(reoffense_rate = mean(two_year_recid, na.rm = TRUE), .groups = "drop")

#not adjusted for the size of each group 
compas %>%
  filter(race == "African-American" | race == "Caucasian") %>%
ggplot(aes(x=nom_decile_score, y = two_year_recid)) +
  geom_col() +
  facet_wrap(~race) +
  labs(title = "Risk Decile Scores and Frequency of Reoffense",
       x="Decile Score",
       y = "# Reoffenders")

#Using proportions of reoffense at each decile 
ggplot(df, mapping = aes(x=nom_decile_score, y=reoffense_rate)) +
  geom_col() +
  facet_wrap(~race) +
  labs(title = "Risk Decile Scores and Proportions of Reoffense",
       x="Decile Score",
       y = "Proportion of Reoffenders")

```

## Exercise 14  

I think the focus should be on optimizing the predictive power of the algorithm for each race. I think fair = accurate predictions, and given how much inequality exists in the justice system, I think its possible that one combination of factors might do a good job predicting for one race, while another combination might be best for another. That said, I don't think its possible to create a fair algorithm, because the data feeding the algorithm is based off of an unjust system. For example, prior offenses are part of the algorithm, but black people are more likely to be unfairly charged for crimes. The justice system underlying the data is unfair, so the algorithm can be as "fair" on paper as we want, but it is based on biased information.

I do think its important to adjust the algorithm so that it is not more or less likely that there is a false positive / false negative based on race - the fact that white people are more likely to get false negatives, and black people false positives, is terrible. 

## Exercise 15 

On the one hand, a fair algorithm might be one that assigns risk based on exactly the same information, and weights this information in the same manner (e.g. prior offenses being weighted the same across races). On the other, a fair algorithm can also be thought of as one that make the most accurate predictions, and this could involve using different combinations of predictors / weighting to accomplish this end. 

## Exercise 16 

The justice system needs to be more fair overall, that way the data the algorithm uses is less biased. I also think that the algorithm should not be interpreted as the "be all end all." There should be multiple other things to consider about the person to make a determination about risk - e.g. face to face interviews, mental health evaluations - to get a more comprehensive understanding of the defendant.  



